Depth over distance: Hawaiian cryptic reef communities are more distinct across ~90 m depth than ~2500 km of shallow habitat

### 2023-10-20 manuscript.Rmd
updates

---
title: "Depth over distance: Hawaiian cryptic reef communities are more distinct across ~90 m depth than ~2500 km of shallow habitat"
author:
  - Kaleonani K.C. Hurley:
      institute:
        - himb
        - nerr
      equal_contributor: yes
      correspondence: yes
      email: "hurleyk@hawaii.edu"
  - Mykle L. Hoban:
      institute:
        - himb
      equal_contributor: yes
  - Molly A. Timmers:
      institute:
        - himb
        - natgeo
  - Kerry Reardon:
      institute: noaa
  - Robert J. Toonen:
      institute: himb
institute:
  - himb: Hawai&#x02BB;i Institute of Marine Biology
  - nerr: He&#x02BB;eia National Estuarine Research Reserve
  - natgeo: National Geographic Pristine Seas
  - noaa: National Oceanic and Atmospheric Administration
abstract:
  Shallow coral reefs are extensively studied, and although scleractinian corals have been recorded to 165 m, little is known about other mesophotic coral reef ecosystem (MCE) inhabitants. Brachyuran crabs fill many ecological and trophic niches on reefs, making them ideal candidates for evaluating species composition among depths. Here we ask if MCEs host the same communities as the shallower reefs spread across the entire 2400 km Hawaiian Archipelago. We deployed Autonomous Reef Monitoring Structures (ARMS) for two years among shallow sites (12 m) across the latitudinal gradient of the Hawaiian Archipelago to compare directly with a depth gradient (12, 30, 60, and 90 m) south of Oʻahu island. A total of 2284 individual crabs representing 90 morphospecies (19 families) were found. Assemblage composition was not statistically significantly different among shallow sites spread across the archipelago but was significantly different and highly stratified across the depth gradient.  We show that 90 m of depth is a more potent determinant of brachyuran community composition than the latitudinal, geographic (high vs low islands), or anthropogenic impact (human population size) gradient across the archipelago. Deeper reefs host significantly different brachyuran communities than shallow ones in Hawai&#x02BB;i, and only 3 of 90 morphospecies (~3%) occurring across the entire depth range (mesophotic to shallow) were sampled. Further research on the cryptic reef-dwelling fauna is needed for identifying areas of unique biodiversity, as well as boundaries between depth-stratified reef ecosystems.
bibliography: references/citations.bib
csl: references/style.csl
output:
  word_document:
    pandoc_args:
      - "--toc-depth=2"
      - "--filter=resources/pandoc-crossref"
      - "--lua-filter=resources/scholarly-metadata.lua"
      - "--lua-filter=resources/author-info-blocks.lua"
      - "--reference-doc=resources/ref.docx"
    keep_md: true
tblPrefix: Table
figPrefix:
  - "figure"
  - "figures"
chapters: True
chaptersDepth: 1
chapDelim: ""
header-includes:
  - \usepackage{bm}
  - \usepackage{upgreek}
---
<!--word count: Proceedings B publishes articles up to 12 pages, including figures and tables, with no more than 10 000 words. The word count includes the title page, author names, abstract, figure legends, tables and the reference list. As a rough guide, 10 000 words = 12 pages without additional figures; 850 words = 1 page and 2-4 small figures = 1 page. For example, an acceptable paper could be made up of 9 000 words and 4 small figures or 8500 words, 2 large figures and 4 small figures.-->
### Keywords {-}
<div custom-style="mono">ARMs, mesophotic coral ecosystems, community structure, dispersal, Hawai&#x02BB;i, brachyuran crabs</div>

@@

The Hawaiian Archipelago comprises high volcanic islands and atolls stretching across ~2,500 km in the North Pacific Ocean, and is typically regarded as two regions: the main Hawaiian Islands (MHI), which consist of the eight inhabited high islands, and the Northwestern Hawaiian Islands (NWHI), which are the uninhabited atolls and banks northwest of Kaua&#x02BB;i and Ni&#x02BB;ihau. The islands are subject to anthropogenic and physical gradients across their span [@selkoeMapHumanImpacts2009], which we hypothesize to influence coral reef communities. Studies examining community structure across environmental gradients have been criticized for confounding multiple potentially covarying factors (such as habitat area and/or complexity) across the measured gradients [e.g., @bellwoodRegionalScaleAssemblyRules2001; @grahamEffectsMarineReserve2003; @connollyCommunityStructureCorals2005; @ewersConfoundingFactorsDetection2006; @fletcherRoleHabitatArea2007a]. Thus, we used autonomous reef monitoring structures (ARMS)---standardized collection devices designed to mimic reef structural complexity---to sample cryptic communities while controlling for habitat<!--differences in habitat variables--> [@zimmermanArtificialReefMatrix2004; @brainardAutonomousReefMonitoring2009]. Specifically, we used ARMS to sample brachyuran crabs, a diverse group of arthropods that comprise a considerable proportion of coral reef cryptofauna [289 species occur in Hawai&#x02BB;i, @castroCatalogAnomuranBrachyuran2011]. We compared crab communities among islands and regions (MHI vs. NWHI) and examined the effects of a suite of possible environmental and anthropogenic drivers. Additionally, given the strong reported differences between deep and shallow reefs across taxa [@hobanPlumbingDepthsEnvironmental2023; @hurleyAssessmentShallowMesophotic2016; @rochaMesophoticCoralEcosystems2018], we also compared the relative contributions of depth (shallow vs. mesophotic) and island region to crab community variation.

Despite conclusions based on intertidal studies, similar generalizations have not emerged to describe coral reef communities [@moraEcologyFishesCoral2015; @sheppardBiologyCoralReefs2017], with existing hypotheses ranging from the neutral theory of biodiversity [@hubbellUnifiedNeutralTheory2011] and lottery recruitment [@saleStockRecruitRelationshipsRegional1982] to competitive networks [@bussCompetitiveNetworksNontransitive1979] and fine-scale resource partitioning driven by predators and herbivores [@abramsRecruitmentLotteriesCoexistence1984; @hixonPredationPreyRefuges1993]. Studies that have investigated these topics on coral reefs have traditionally focused on conspicuous taxa such as fishes and corals [@cite], yet these groups constitute a minority of total reef biodiversity [@cite]. The majority of coral reef biodiversity is cryptic, inhabiting interstitial spaces, and little is known about factors contributing to     Zooxanthellate corals are the ecosystem engineers of reef habitats but are restricted to specific areas characterized by a relatively narrow range of chemical and physical parameters<!-- (i.e., tropical and subtropical warm waters, oligotrophic conditions, 32-40 PSU salinity, and sufficient incident light, which decreases rapidly below ~30 m depth)-->. Shallow coral reef ecosystems cover less than 0.2% of the seafloor and in such a small area they hold the 32-38% of all marine species [@fisherSpeciesRichnessCoral2015]. Yet the <!--drivers of high biodiversity and the -->mechanisms by which communities are structured on coral reefs remain a subject of debate [@carrBiodiversityPopulationRegulation2002; @connollyIndoPacificBiodiversityCoral2003; @almanyDoesIncreasedHabitat2004; @bellwoodEnvironmentalGeometricConstraints2005, 2006; @dornelasCoralReefDiversity2006; @grahamImportanceStructuralComplexity2013; @komyakovaRelativeImportanceCoral2013]. *Add some text about community structuring across space, rather than just at smaller scales*.

A<!--nother--> likely important but under-investigated aspect of coral reef community structure is depth. Most studies to date have focused on the upper ~30 m of the water column, yet coral reefs and associated communities are known to extend beyond 150 m [@pyleExploringDeepCoral1996; @pyleMesophoticCoralEcosystems2019], with scleractinian corals reported to 170 m [@rouzeSymbioticAssociationsDeepest2021]. As much as two thirds of coral reef ecosystems lie in this under-explored zone [@hindersteinThemeSectionMesophotic2010]. These habitats, known as mesophotic coral ecosystems (MCEs), are typically defined as those extending from ~30--150 m depth [@hindersteinThemeSectionMesophotic2010; @kahngCommunityEcologyMesophotic2010]. <!--and were believed to experience dampened disturbances in comparison to shallow reefs and could therefore act as deep water refugia for shallow reef species threatened by anthropogenic impacts --> They have been hypothesized as possible refugia from disturbances on shallow reefs [@bongaertsAssessingDeepReef2010; @smithDepthRefugiumCatastrophic2014; @holsteinFertileFathomsDeep2015; @muirSpeciesIdentityDepth2017], although a number of studies have found evidence that deep reefs are unique and uniquely threatened, challenging this notion [@hobanPlumbingDepthsEnvironmental2023; @hurleyAssessmentShallowMesophotic2016; @kaneHighLevelsMesophotic2014; @pinheiroDeepReefFishes2019; @rochaMesophoticCoralEcosystems2018]. <!--However, recent literature describing species compositions along depth gradients reveal a more complex picture among taxa and locations because some groups may use mesophotic reefs as a refuge in some places, while others show community turnover with depth such that MCE habitats are ecologically distinct from adjacent shallow water coral reefs [@hurleyAssessmentShallowMesophotic2016; @smithCaribbeanMesophoticCoral2016; @semmlerAreMesophoticCoral2017; @bongaertsDeepReefsAre2017; @rochaMesophoticCoralEcosystems2018]. Further, there are a range of connections depending on the taxa – large, mobile organisms like fish tend to have more shared species across depth than sedentary organisms such as corals and algae, but even among the fishes the community structure of MCEs is often distinct [@kaneHighLevelsMesophotic2014; @pyleComprehensiveInvestigationMesophotic2016 ; @asherMesophoticDepthGradients2017; @fukunagaChangesMesophoticReef2017; @colemanShiftingReefFish2018; @pinheiroDeepReefFishes2019]. Although the majority of coral reef biodiversity consists of the algal and invertebrate cryptobenthic organisms living on and within the reef structure [@fautinOverviewMarineBiodiversity2010; @knowltonCoralReefBiodiversity2010; @fisherSpeciesRichnessCoral2015], the studies listed above focus almost exclusively on the corals and fishes which are more easily surveyed.-->

Here, we employed autonomous reef monitoring structures (ARMS)  Using ARMS, we sampled brachyuran crab communities across the Hawaiian Archipelago to test the effects of environmental and anthropogenic drivers on the structure of cryptic reef communities.    Given the number and magnitude of differences between these regions, we hypothesized that species diversity would be different between regions (MHI vs NWHI). We therefore looked at community composition along the latitudinal gradient and investigated community composition in response to physical gradients (depth, temperature), biological gradients (coral cover, chlorophyll concentration), geological gradients (island slope, distances between neighboring islands), and anthropogenic gradients (human population count, human impacts). We hypothesized that the communities found across the latitudinal gradient encompassing all of these environmental variables would show fewer differences between individual islands or atolls, but would be greatest between the two regions, MHI and NWHI.  Finally, to investigate whether a latitudinal gradient or depth gradient would be a stronger force in driving community composition, we compared the reef crab assemblage from across the shallow reefs in the Hawaiian Archipelago to reef crab assemblages across a mesophotic depth gradient.

The Hawaiian Archipelago is a volcanic island arc made up of high islands and atolls spanning ~9 degrees of latitude and ~2,500 km. It is typically regarded as two regions: the inhabited main Hawaiian Islands (MHI), which include the eight islands from Hawai&#x02BB;i Island to Kauai&#x02BB;i, and the largely uninhabited Northwestern Hawaiian Islands (NWHI), which constitute the 10 low islands and atolls to the northwest of Kaua&#x02BB;i. We compared shallow reef crab communities across the 2500 km archipelago and  we compared reef crab assemblages from the shallow reefs to the reef crab assemblages across a mesophotic depth gradient off of Kewalo, Oʻahu. By controlling for habitat area and complexity and by incorporating sites across various environmental and human impact gradients, we sought to better understand whether latitude, the environment, or depth is more influential in structuring reef crab communities.

@@

```{r fig-map, fig.cap='Map of ARMS deployment site across the Hawaiian Archipelago (Papahānaumokuākea Marine National Monument is shaded in light blue)', results="asis", fig.width=9, fig.height=6, dependson="init"}
library(sf)
library(rnaturalearth)

@@

ARMS units were retrieved by encapsulating them in-situ in a mesh-lined crate (100 μm pore size) to prevent escape of motile organisms. Once returned to the surface, they were placed in containers of aerated, filtered seawater, crates were removed, and units were systematically disassembled plate-by-plate. Seawater surrounding disassembled units was sieved through 2 mm, 500 μm, and 100 μm mesh-size geologic sieves [see @lerayDNABarcodingMetabarcoding2015 for processing details], and retained macroorganisms were preserved in 95% ethanol. All adult brachyuran crabs ($\geq$ 5 mm) were photographed, identified to morphospecies, and preserved in 95% ethanol. Individual preserved specimens were accessioned into the invertebrate zoology collections at the Florida Museum of Natural History (catalog numbers xxx–xxx).

@@

We used distance-based redundancy analysis (dbRDA; a form of multivariate multiple regression) with `dbrda` to examine the relative contribution of environmental variables to crab community variation across the islands. We first used a stepwise forward selection procedure (`ordiR2step`) to find the variables that best explained community dissimilarity. Then, we sequentially removed variables with the largest variance inflation factors (VIF; calculated using `vif.cca`) until all values were <3, indicating variables were sufficiently uncorrelated. Model significance was assessed using permutation tests with `r num(PERM,1)` permutations (function `anova`). Continuous variables were scaled to unit variance. We assessed variation among groups using `betadisper` and `adonis2`. These functions implement analysis of multivariate homogeneity of group dispersions (PERMDISP) and multivariate permutational analysis of variance (PERMANOVA) respectively. PERMDISP is used to assess variability within groups and PERMANOVA is used to assess differences among groups. Significance was assessed with permutation tests in both cases. Variation among sampling groups (shallow MHI, shallow NWHI, O&#x02BB;ahu 30 m, 60 m, 90 m) was further examined and visualized using canonical analysis of principal coordinates (CAP) based on discriminant analysis [@andersonCanonicalAnalysisPrincipal2003] as implemented in the `CAPdiscrim` function in `BiodiversityR` [@kindtTreeDiversityAnalysis2005].

@@

Potential environmental drivers of shallow crab community structure were assembled from published literature [@selkoeDNACoralReef2016], NOAA's National Coral Reef Monitoring Program (NCRMP), and the Pacific Islands Ocean Observing System (PacIOOS) ([@tbl:env_table]). Due to logistical limitations, environmental factors are only available for shallow samples.

@@

The `r nsamples(cc$crabs_untransformed)` ARMS units deployed across the Hawaiian Islands (in `r format_range(cc$crab_data$depth)` m depth) yielded a total of `r num(sum(sample_sums(cc$crabs_untransformed)),1)` adult crabs. Units contained `r num(all_summary$abundance_mean,0.1)` $\pm$ `r num(all_summary$abundance_sd,0.1)` (mean $\pm$ SD) individuals, and abundance ranged from `r format_range(all_summary$abundance_range)`. Crabs were identified to `r sum(unlist(all_summary$types_breakdown))` distinct types (including `r all_summary$types_breakdown$species` species and `r all_summary$types_breakdown$morphospecies` morphospecies) in `r all_summary$taxonomy_breakdown$genera` genera and `r all_summary$taxonomy_breakdown$families` families, which accounted for ~36% of the known crab diversity reported from Hawaiian waters [@castroCatalogAnomuranBrachyuran2011]. The four most abundant species were: `r with(all_summary,top_list(top,top_n,top_units,fig="fig_crabs"))`. `r str_to_sentence(english(all_summary$singleton_count))` species occurred in a single unit across all sites (singletons), accounting for ~`r num(all_summary$singleton_count/ntaxa(cc$crabs_untransformed)*100,1)`% of total recovered crab species.

@@

Shallow crab community structure was significantly different between island groups (MHI vs. NWHI) (`r permanova(anova_region,0.01)`) as well as among individual islands (`r permanova(anova_island,0.01)`; see [@fig:fig_pairwise] for results of pairwise comparisons). There was no significant difference in community dispersion between the MHI and NWHI (`r permdisp(bd_region)`). When mesophotic ($\geq$ 30 m) ARMS units were included, there were significant differences among depth zones (\< 30 m, 30 m, 60 m, 90 m) (`r permanova(anova_depth,0.01)`) and community dispersion (`r permdisp(bd_depth)`). However, community dispersion was not different among depth zones $\geq$ 30 m (`r permdisp(bd_deep)`). Although crab communities varied significantly among both depth zones and island groups when examining the complete dataset (`r permanova(anova_both,0.01)`), depth explained `r num(effect_ratio,0.1)` times more variability than island group (based on model $\mathrm{R}^2$). In the CAP analysis, ARMS units were correctly assigned to sampling groups `r pc(cap_depth$percent/100,0.1)` of the time. The first two axes explained `r pc(cap_var$x,0.1)` and `r pc(cap_var$y,0.1)` of the variation, respectively. The $\geq$ 30 m samples from O&#x02BB;ahu grouped separately from the shallow samples collected across the entire ~2,500 km of the Hawaiian Islands ([@fig:fig_cap]). In addition, there was as much variability across the 30--90 m samples as across all shallow samples, with the 30 m samples spread out primarily along the LD2 (y) axis and the 60 m and 90 m samples forming a tight group ([@fig:fig_cap]).

@@

```{r fig-alpha, fig.cap='Box plots of Simpson diversity by (A) island group and (B) sampling depth for complete dataset (all depths). Notches represent nonparametric confidence intervals about the median', results="asis", dev="svg", fig.width=10, fig.height=10, dependson="init"}
library(patchwork)

@@

```{r fig-alpha2, fig.cap='Box plots of (A) species richness and (B) Simpson diversity and by island group for shallow samples only.', results="asis", dev="svg", fig.width=8, fig.height=4, dependson="init"}
art <- shallow_alpha$richness_table %>% 
  mutate(island_group = str_to_sentence(island_group)) %>%
  mutate(shallow_deep = str_to_sentence(shallow_deep),shallow_deep = fct_reorder(shallow_deep,depth))

@@

```{r fig-pca, fig.cap='Principal component analysis (PCA) of environmental variables by sampling site. Ellipses represent 95% confidence intervals around island groups', results='asis', dev='svg', fig.width=8,fig.height=8, dependson='init'}
cds <- cc$crab_data_shallow

@@

```{r pairwise-island, fig.cap='Significance levels of pairwise PERMANOVA tests by island (FDR-adjusted p-values)', results="asis", dev="svg", fig.width=12, fig.height=9, dependson="init"}
library(ggtext)
pairwise_island <- pairwise_adonis(cc$crab_dist_shallow,cc$crab_data_shallow$island,permutations=PERM)

>>>>

---
title: "Depth over distance: Hawaiian cryptic reef communities are more distinct across ~90 m depth than ~2500 km of shallow habitat"
author:
  - Kaleonani K.C. Hurley:
      institute:
        - himb
        - nerr
      equal_contributor: yes
      correspondence: yes
      email: "hurleyk@hawaii.edu"
  - Mykle L. Hoban:
      institute:
        - himb
      equal_contributor: yes
  - Molly A. Timmers:
      institute:
        - himb
        - natgeo
  - Kerry Reardon:
      institute: noaa
  - Robert J. Toonen:
      institute: himb
institute:
  - himb: Hawai&#x02BB;i Institute of Marine Biology
  - nerr: He&#x02BB;eia National Estuarine Research Reserve
  - natgeo: National Geographic Pristine Seas
  - noaa: National Oceanic and Atmospheric Administration
abstract:
  Shallow coral reefs are extensively studied, and although scleractinian corals have been recorded to 165 m, little is known about other mesophotic coral reef ecosystem (MCE) inhabitants. Brachyuran crabs fill many ecological and trophic niches on reefs, making them ideal candidates for evaluating species composition among depths. Here we ask if MCEs host the same communities as the shallower reefs spread across the entire 2400 km Hawaiian Archipelago. We deployed Autonomous Reef Monitoring Structures (ARMS) for two years among shallow sites (12 m) across the latitudinal gradient of the Hawaiian Archipelago to compare directly with a depth gradient (12, 30, 60, and 90 m) south of Oʻahu island. A total of 2284 individual crabs representing 90 morphospecies (19 families) were found. Assemblage composition was not statistically significantly different among shallow sites spread across the archipelago but was significantly different and highly stratified across the depth gradient.  We show that 90 m of depth is a more potent determinant of brachyuran community composition than the latitudinal, geographic (high vs low islands), or anthropogenic impact (human population size) gradient across the archipelago. Deeper reefs host significantly different brachyuran communities than shallow ones in Hawai&#x02BB;i, and only 3 of 90 morphospecies (~3%) occurring across the entire depth range (mesophotic to shallow) were sampled. Further research on the cryptic reef-dwelling fauna is needed for identifying areas of unique biodiversity, as well as boundaries between depth-stratified reef ecosystems.
bibliography: references/citations.bib
csl: references/style.csl
output:
  word_document:
    pandoc_args:
      - "--toc-depth=2"
      - "--filter=resources/pandoc-crossref"
      - "--lua-filter=resources/scholarly-metadata.lua"
      - "--lua-filter=resources/author-info-blocks.lua"
      - "--reference-doc=resources/ref.docx"
    keep_md: true
tblPrefix: table
figPrefix:
  - "figure"
  - "figures"
chapters: True
chaptersDepth: 1
chapDelim: ""
header-includes:
  - \usepackage{bm}
  - \usepackage{upgreek}
---
<!--word count: Proceedings B publishes articles up to 12 pages, including figures and tables, with no more than 10 000 words. The word count includes the title page, author names, abstract, figure legends, tables and the reference list. As a rough guide, 10 000 words = 12 pages without additional figures; 850 words = 1 page and 2-4 small figures = 1 page. For example, an acceptable paper could be made up of 9 000 words and 4 small figures or 8500 words, 2 large figures and 4 small figures.-->
### Keywords {-}
<div custom-style="mono">ARMs, mesophotic coral ecosystems, community structure, dispersal, Hawai&#x02BB;i, brachyuran crabs</div>

@@

The Hawaiian Archipelago comprises high volcanic islands and atolls stretching across ~2,500 km in the North Pacific Ocean, and is typically regarded as two regions: the main Hawaiian Islands (MHI), which include the eight inhabited high islands, and the Northwestern Hawaiian Islands (NWHI), the uninhabited atolls and banks northwest of Kaua&#x02BB;i and Ni&#x02BB;ihau. The islands are subject to anthropogenic and physical gradients across their span [@selkoeMapHumanImpacts2009], which we hypothesize to influence coral reef communities. Studies examining community structure across environmental gradients have been criticized for confounding multiple potentially covarying factors (such as habitat area and/or complexity) across the measured gradients [e.g., @bellwoodRegionalScaleAssemblyRules2001; @grahamEffectsMarineReserve2003; @connollyCommunityStructureCorals2005; @ewersConfoundingFactorsDetection2006; @fletcherRoleHabitatArea2007a]. Thus, we used autonomous reef monitoring structures (ARMS)---standardized collection devices designed to mimic reef structural complexity---to sample cryptic communities while controlling for habitat<!--differences in habitat variables--> [@zimmermanArtificialReefMatrix2004; @brainardAutonomousReefMonitoring2009]. Specifically, we used ARMS to sample brachyuran crabs, a group comprising a notable proportion of the coral reef cryptofauna [289 species in Hawai&#x02BB;i, @castroCatalogAnomuranBrachyuran2011]. We compared crab communities among islands and regions (MHI vs. NWHI) and examined the effects of a suite of possible environmental and anthropogenic drivers.

A likely important but under-investigated aspect of coral reef community structure is depth. Most studies to date have focused on the upper ~30 m of the water column, yet coral reefs and associated communities are known to extend beyond 150 m [@pyleExploringDeepCoral1996; @pyleMesophoticCoralEcosystems2019], with scleractinian corals reported to 170 m [@rouzeSymbioticAssociationsDeepest2021]. As much as two thirds of coral reef ecosystems lie in this under-explored zone [@hindersteinThemeSectionMesophotic2010]. These habitats, known as mesophotic coral ecosystems (MCEs), are typically defined as those extending from ~30--150 m depth [@hindersteinThemeSectionMesophotic2010; @kahngCommunityEcologyMesophotic2010]. <!--and were believed to experience dampened disturbances in comparison to shallow reefs and could therefore act as deep water refugia for shallow reef species threatened by anthropogenic impacts --> They have been hypothesized as possible refugia from disturbances on shallow reefs [@bongaertsAssessingDeepReef2010; @smithDepthRefugiumCatastrophic2014; @holsteinFertileFathomsDeep2015; @muirSpeciesIdentityDepth2017], although a number of studies have found evidence that deep reefs are unique and uniquely threatened, challenging this notion [@diazLightTemperatureDrive; @diazMesophoticCoralBleaching2023; @hobanPlumbingDepthsEnvironmental2023; @hurleyAssessmentShallowMesophotic2016; @kaneHighLevelsMesophotic2014; @pinheiroDeepReefFishes2019; @rochaMesophoticCoralEcosystems2018]. <!--However, recent literature describing species compositions along depth gradients reveal a more complex picture among taxa and locations because some groups may use mesophotic reefs as a refuge in some places, while others show community turnover with depth such that MCE habitats are ecologically distinct from adjacent shallow water coral reefs [@hurleyAssessmentShallowMesophotic2016; @smithCaribbeanMesophoticCoral2016; @semmlerAreMesophoticCoral2017; @bongaertsDeepReefsAre2017; @rochaMesophoticCoralEcosystems2018]. Further, there are a range of connections depending on the taxa – large, mobile organisms like fish tend to have more shared species across depth than sedentary organisms such as corals and algae, but even among the fishes the community structure of MCEs is often distinct [@kaneHighLevelsMesophotic2014; @pyleComprehensiveInvestigationMesophotic2016 ; @asherMesophoticDepthGradients2017; @fukunagaChangesMesophoticReef2017; @colemanShiftingReefFish2018; @pinheiroDeepReefFishes2019]. Although the majority of coral reef biodiversity consists of the algal and invertebrate cryptobenthic organisms living on and within the reef structure [@fautinOverviewMarineBiodiversity2010; @knowltonCoralReefBiodiversity2010; @fisherSpeciesRichnessCoral2015], the studies listed above focus almost exclusively on the corals and fishes which are more easily surveyed.--> Given the strong reported differences between deep and shallow reefs across taxa and the paucity of research on drivers of cryptic coral reef community variation, we sought to answer the following research questions: (1) How do shallow crab communities vary along the natural and anthropogenic gradients of the Hawaiian Archipelago? (2) Do shallow crab communities differ between the Main and Northwestern Hawaiian Islands? (3) What environmental and/or anthropogenic factors drive crab community structure in shallow coral reefs across the Hawaiian Archipelago? (4) What are the relative contributions of depth (shallow vs. mesophotic) and island region to crab community variation across the Hawaiian Archipelago?

<!-- Here, we employed autonomous reef monitoring structures (ARMS)  Using ARMS, we sampled brachyuran crab communities across the Hawaiian Archipelago to test the effects of environmental and anthropogenic drivers on the structure of cryptic reef communities.    Given the number and magnitude of differences between these regions, we hypothesized that species diversity would be different between regions (MHI vs NWHI). We therefore looked at community composition along the latitudinal gradient and investigated community composition in response to physical gradients (depth, temperature), biological gradients (coral cover, chlorophyll concentration), geological gradients (island slope, distances between neighboring islands), and anthropogenic gradients (human population count, human impacts). We hypothesized that the communities found across the latitudinal gradient encompassing all of these environmental variables would show fewer differences between individual islands or atolls, but would be greatest between the two regions, MHI and NWHI.  Finally, to investigate whether a latitudinal gradient or depth gradient would be a stronger force in driving community composition, we compared the reef crab assemblage from across the shallow reefs in the Hawaiian Archipelago to reef crab assemblages across a mesophotic depth gradient.  -->

<!-- The Hawaiian Archipelago is a volcanic island arc made up of high islands and atolls spanning ~9 degrees of latitude and ~2,500 km. It is typically regarded as two regions: the inhabited main Hawaiian Islands (MHI), which include the eight islands from Hawai&#x02BB;i Island to Kauai&#x02BB;i, and the largely uninhabited Northwestern Hawaiian Islands (NWHI), which constitute the 10 low islands and atolls to the northwest of Kaua&#x02BB;i. We compared shallow reef crab communities across the 2500 km archipelago and  we compared reef crab assemblages from the shallow reefs to the reef crab assemblages across a mesophotic depth gradient off of Kewalo, Oʻahu. By controlling for habitat area and complexity and by incorporating sites across various environmental and human impact gradients, we sought to better understand whether latitude, the environment, or depth is more influential in structuring reef crab communities. -->

@@

```{r fig-map, fig.cap='Map of ARMS deployment sites across the Hawaiian Archipelago (Papahānaumokuākea Marine National Monument is shaded in light blue).', results="asis", fig.width=9, fig.height=6, dependson="init"}
library(sf)
library(rnaturalearth)

@@

ARMS units were retrieved by encapsulating them in-situ in a mesh-lined crate (100 μm pore size) to prevent escape of motile organisms. Once returned to the surface, they were placed in containers of aerated, filtered seawater, crates were removed, and units were systematically disassembled plate-by-plate. Seawater surrounding disassembled units was sieved through 2 mm, 500 μm, and 100 μm mesh-size geologic sieves [see @lerayDNABarcodingMetabarcoding2015 for processing details], and retained macroorganisms were preserved in 95% ethanol. All adult brachyuran crabs ($\geq$ 5 mm) were photographed, identified to morphospecies, and preserved in 95% ethanol. Individual preserved specimens were accessioned into the invertebrate zoology collections at the Florida Museum of Natural History ([catalog numbers xxx--xxx]{.mark}).

@@

We used distance-based redundancy analysis (dbRDA; a form of multivariate multiple regression) with `dbrda` to examine the relative contribution of environmental variables to crab community variation across the islands. We first used a stepwise forward selection procedure (`ordiR2step`) to find the variables that best explained community dissimilarity. Then, we sequentially removed variables with the largest variance inflation factors (VIF; calculated using `vif.cca`) until all values were <3, indicating variables were sufficiently uncorrelated. Model significance was assessed using permutation tests with `r num(PERM,1)` permutations in the function `anova`. Continuous variables were scaled to unit variance. We assessed variation among groups using `betadisper` and `adonis2`. These functions implement analysis of multivariate homogeneity of group dispersions (PERMDISP) and multivariate permutational analysis of variance (PERMANOVA) respectively. PERMDISP is used to assess variability within groups and PERMANOVA is used to assess differences among groups. Significance was assessed with permutation tests in both cases. Variation among sampling groups (shallow MHI, shallow NWHI, O&#x02BB;ahu 30 m, 60 m, 90 m) was further examined and visualized using canonical analysis of principal coordinates (CAP) based on discriminant analysis [@andersonCanonicalAnalysisPrincipal2003] as implemented in the `CAPdiscrim` function in `BiodiversityR` [@kindtTreeDiversityAnalysis2005].

@@

Potential environmental drivers of shallow crab community structure were assembled from published literature [@selkoeDNACoralReef2016], NOAA's National Coral Reef Monitoring Program (NCRMP), and the Pacific Islands Ocean Observing System (PacIOOS) ([@tbl:env_table]). Due to logistical limitations, environmental factors were only available for shallow samples.

@@

The `r nsamples(cc$crabs_untransformed)` ARMS units deployed across the Hawaiian Islands (in `r format_range(cc$crab_data$depth)` m depth) yielded a total of `r num(sum(sample_sums(cc$crabs_untransformed)),1)` adult crabs. Units contained `r num(all_summary$abundance_mean,0.1)` $\pm$ `r num(all_summary$abundance_sd,0.1)` (mean $\pm$ SD) individuals, and abundance ranged from `r format_range(all_summary$abundance_range)`. Crabs were identified to `r sum(unlist(all_summary$types_breakdown))` distinct types (`r all_summary$types_breakdown$species` species and `r all_summary$types_breakdown$morphospecies` morphospecies) in `r all_summary$taxonomy_breakdown$genera` genera and `r all_summary$taxonomy_breakdown$families` families, which accounted for ~36% of the known crab diversity reported from Hawaiian waters [@castroCatalogAnomuranBrachyuran2011]. The four most abundant species were: `r with(all_summary,top_list(top,top_n,top_units,fig="fig_crabs"))`. `r str_to_sentence(english(all_summary$singleton_count))` species occurred in a single unit across all sites (singletons), accounting for ~`r num(all_summary$singleton_count/ntaxa(cc$crabs_untransformed)*100,1)`% of total recovered crab species.

@@

Shallow crab community structure was significantly different between island groups (MHI vs. NWHI) (`r permanova(anova_region,0.01)`) as well as among individual islands (`r permanova(anova_island,0.01)`; see [@fig:fig_pairwise] for results of pairwise comparisons). There was no significant difference in community dispersion between the MHI and NWHI (`r permdisp(bd_region)`). When mesophotic ($\geq$ 30 m) ARMS units were included, there were significant differences among depth zones (\< 30 m, 30 m, 60 m, 90 m) (`r permanova(anova_depth,0.01)`) and community dispersion (`r permdisp(bd_depth)`). However, community dispersion was not different among depth zones $\geq$ 30 m (`r permdisp(bd_deep)`). Although crab communities varied significantly among both depth zones and island groups when examining the complete dataset (`r permanova(anova_both,0.01)`), depth explained `r num(effect_ratio,0.1)` times more variability than island group (based on model $\mathrm{R}^2$). In the CAP analysis, ARMS units were correctly assigned to sampling groups `r pc(cap_depth$percent/100,0.1)` of the time. The first two linear discriminant axes explained `r pc(cap_var$x,0.1)` and `r pc(cap_var$y,0.1)` of the variation, respectively. The mesophotic samples ($\geq$ 30 m) from O&#x02BB;ahu grouped separately from the shallow samples collected across the entire ~2,500 km of the Hawaiian Islands ([@fig:fig_cap]). In addition, there was as much variability across the 30--90 m samples as across all shallow samples, with the 30 m samples spread along the LD2 (vertical) axis and the 60 m and 90 m samples grouped together ([@fig:fig_cap]).

@@

```{r fig-alpha, fig.cap='Box plots of (A,B) species richness and (C,D) Simpson diversity for complete dataset (all sampled depths). (A) Species richness by island group. (B) Species richness by sampling depth. (C) Simpson diversity by island group. (D) Simpson diversity by sampling depth. Notches represent nonparametric confidence intervals about the median.', results="asis", dev="svg", fig.width=10, fig.height=10, dependson="init"}
library(patchwork)

@@

```{r fig-alpha2, fig.cap='Box plots of (A) species richness and (B) Simpson diversity by island group for shallow samples only. Notches as in [@fig:fig_alpha].', results="asis", dev="svg", fig.width=8, fig.height=4, dependson="init"}
art <- shallow_alpha$richness_table %>% 
  mutate(island_group = str_to_sentence(island_group)) %>%
  mutate(shallow_deep = str_to_sentence(shallow_deep),shallow_deep = fct_reorder(shallow_deep,depth))

@@

```{r fig-pca, fig.cap='Principal component analysis (PCA) of environmental variables by sampling site. Ellipses represent 95% confidence intervals around island groups.', results='asis', dev='svg', fig.width=8,fig.height=8, dependson='init'}
cds <- cc$crab_data_shallow

@@

```{r pairwise-island, fig.cap='Significance levels of pairwise PERMANOVA tests by island for shallow crab communities (p-values adjusted according to the false discovery rate).', results="asis", dev="svg", fig.width=12, fig.height=9, dependson="init"}
library(ggtext)
pairwise_island <- pairwise_adonis(cc$crab_dist_shallow,cc$crab_data_shallow$island,permutations=PERM)
